---
title: 'Cyclistic Case Study: Bike Sharing'
author: "Nadario Ferguson"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(dplyr)
library(tidyr)
library(geosphere)
```

## Step 1: Collect The Data

```{r}
March_Trips21 <- read_csv("/Users/nadarioferguson/Desktop/Data Analytics Case Study/Case Study 1 - Cyclistic/Dataset Information/202103-divvy-tripdata.csv")
April_Trips21 <- read_csv("/Users/nadarioferguson/Desktop/Data Analytics Case Study/Case Study 1 - Cyclistic/Dataset Information/202104-divvy-tripdata.csv")
May_Trips21 <- read_csv("/Users/nadarioferguson/Desktop/Data Analytics Case Study/Case Study 1 - Cyclistic/Dataset Information/202105-divvy-tripdata.csv")
June_Trips21 <- read_csv("/Users/nadarioferguson/Desktop/Data Analytics Case Study/Case Study 1 - Cyclistic/Dataset Information/202106-divvy-tripdata.csv")
July_Trips21 <- read_csv("/Users/nadarioferguson/Desktop/Data Analytics Case Study/Case Study 1 - Cyclistic/Dataset Information/202107-divvy-tripdata.csv")
August_Trips21 <- read_csv("/Users/nadarioferguson/Desktop/Data Analytics Case Study/Case Study 1 - Cyclistic/Dataset Information/202108-divvy-tripdata.csv")
September_Trips21 <- read_csv("/Users/nadarioferguson/Desktop/Data Analytics Case Study/Case Study 1 - Cyclistic/Dataset Information/202109-divvy-tripdata.csv")
October_Trips21 <- read_csv("/Users/nadarioferguson/Desktop/Data Analytics Case Study/Case Study 1 - Cyclistic/Dataset Information/202110-divvy-tripdata.csv")
November_Trips21 <- read_csv("/Users/nadarioferguson/Desktop/Data Analytics Case Study/Case Study 1 - Cyclistic/Dataset Information/202111-divvy-tripdata.csv")
December_Trips21 <- read_csv("/Users/nadarioferguson/Desktop/Data Analytics Case Study/Case Study 1 - Cyclistic/Dataset Information/202112-divvy-tripdata.csv")
January_Trips22 <- read_csv("/Users/nadarioferguson/Desktop/Data Analytics Case Study/Case Study 1 - Cyclistic/Dataset Information/202201-divvy-tripdata.csv")
February_Trips22 <- read_csv("/Users/nadarioferguson/Desktop/Data Analytics Case Study/Case Study 1 - Cyclistic/Dataset Information/202202-divvy-tripdata.csv")
```
## Convert ride_id and rideable_type to character to be used correctly

```{r}
March_Trips21 <- mutate(March_Trips21, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
April_Trips21 <- mutate(April_Trips21, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
May_Trips21 <- mutate(May_Trips21, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
June_Trips21 <- mutate(June_Trips21, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
July_Trips21 <- mutate(July_Trips21, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
August_Trips21 <- mutate(August_Trips21, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
September_Trips21 <- mutate(September_Trips21, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
October_Trips21 <- mutate(October_Trips21, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
November_Trips21 <- mutate(November_Trips21, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
December_Trips21 <- mutate(December_Trips21, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
January_Trips22 <- mutate(January_Trips22, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
February_Trips22 <- mutate(February_Trips22, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
```

## Combine all dataframes together into 1

```{r}
All_Tripsv1 <- bind_rows(March_Trips21, April_Trips21, May_Trips21, June_Trips21, July_Trips21, August_Trips21, September_Trips21, October_Trips21, November_Trips21, December_Trips21, January_Trips22, February_Trips22)
```

## Add ride_length calcualtions to Dataframe

```{r}
All_Tripsv1$ride_length <- difftime(All_Tripsv1$ended_at, All_Tripsv1$started_at)
```

## Organize the data format and  add a ride_length column

```{r}
All_Tripsv1$date <- as.Date(All_Tripsv1$started_at)
All_Tripsv1$month <- format(as.Date(All_Tripsv1$date), "%m")
All_Tripsv1$day <- format(as.Date(All_Tripsv1$date), "%d")
All_Tripsv1$year <- format(as.Date(All_Tripsv1$date), "%Y")
All_Tripsv1$day_of_week <- format(as.Date(All_Tripsv1$date), "%A")
```

## Check for NA's

```{r}
sum(is.na(All_Tripsv1))
```

## Remove NA's

```{r}
All_Trips_Clean <- na.omit(All_Tripsv1)
```

## Convert ridelength to hour

```{r}
All_Trips_Clean$ride_length_hr <- round(All_Trips_Clean$ride_length/3600, 2)
```

## Convert ride_length to numeric value

```{r}
All_Trips_Clean$ride_length <- as.numeric(as.character(All_Trips_Clean$ride_length))
is.numeric(All_Trips_Clean$ride_length)
```
## Calculate distance

```{r}
All_Trips_Clean$distHaversine <- round(distHaversine(All_Trips_Clean[,c("start_lng", "start_lat")], All_Trips_Clean[,c("end_lng", "end_lat")])/1609.344, 3)
```

## Reassign labels

```{r}
All_Trips_Clean <- All_Trips_Clean %>% 
  mutate(member_casual = recode(member_casual, "Subscriber" = "member", "Customer" = "Casual"))
```

## Removing negative values to create a cleaner dataframe

```{r}
All_Trips_Clean <- All_Trips_Clean[!(All_Trips_Clean$ride_length < 0),]


```

#Phase 4: Analysis

Calculating Statistical Essentials

Calculate the mean, median, maximum and minimum values of cleaned DF

```{r}
mean(All_Trips_Clean$ride_length)
median(All_Trips_Clean$ride_length)
max(All_Trips_Clean$ride_length)
min(All_Trips_Clean$ride_length)
```

Data Aggregation - Compare the member and casual users

```{r}
aggregate(All_Trips_Clean$ride_length ~ All_Trips_Clean$member_casual, FUN = mean)
aggregate(All_Trips_Clean$ride_length ~ All_Trips_Clean$member_casual, FUN = median)
aggregate(All_Trips_Clean$ride_length ~ All_Trips_Clean$member_casual, FUN = max)
aggregate(All_Trips_Clean$ride_length ~ All_Trips_Clean$member_casual, FUN = min)
```

Calculate the average ride time by each day for members vs casual users

```{r}
aggregate(All_Trips_Clean$ride_length ~ All_Trips_Clean$member_casual + All_Trips_Clean$day_of_week, FUN = mean)
```

Calculate the average distance by each day for members vs casual users

```{r}
aggregate(All_Trips_Clean$distHaversine ~ All_Trips_Clean$member_casual + All_Trips_Clean$day_of_week, FUN = mean)
```


Analysis

Organize the days of the week

```{r}
All_Trips_Clean$day_of_week <- ordered(All_Trips_Clean$day_of_week, levels = c("Sunday", "Monday", "Tuesday", "Wedneday", "Thursday", "Friday", "Saturday"))
```


Analyze the bike-share data by type and weekday

```{r}
All_Trips_Clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length), average_distance = mean(distHaversine)) %>% 
  arrange(member_casual, weekday)
```

##Phase 5: Share

Visualizing the number of rides by rider type

```{r}
All_Trips_Clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length), average_distance = mean(distHaversine)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) + geom_col(position = "dodge")
```
Create a visualization for average duration

```{r}
All_Trips_Clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) + geom_col(position = "dodge")
```

Create a visualization for average duration

```{r}
All_Trips_Clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_distance = mean(distHaversine)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = average_distance, fill = member_casual)) + geom_col(position = "dodge")
```

Duration vs Distrance Line Graph for member vs casual users

```{r}
ggplot(filter(All_Trips_Clean, All_Trips_Clean$ride_length<600)) +
  geom_histogram(mapping = aes(x = ride_length)) +
  facet_wrap(~member_casual)
```

Phase 6: Act

During this phase we methodically go through our data problem and the analysis we made to make a data-driven decision.

What does the data tells us?

We can clearly see that membership users are consistent throughout the weekdays and then decline during the weekend. This is the complete opposite for casual users. Additionally, casual users have a much higher duration time than members. With this, my recommendation to my manager, would be to introduce a price structure that placese emphasis on duration vs the number of rides. The pricing structure should reward membership with a longer allotted time.This would incentivize casual users to purchase memberships if it rewarded them with a longer time to use the bike for a cheaper cost than not being within the membership.


